---
title: "Exercise prediction and Quality Analysis"
author: "Danish Tamboli"
date: "9/2/2020"
output:
  html_document: default
  pdf_document: default
---

### Background:
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks.

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants. 

They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

### Dataset Download:

Training Data is available here: [Training Data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)  

Testing Data is available here: [Testing Data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

```{r file check/ download}

# Checking if given training set data is locally available, else to download.
if(!file.exists("pml-training.csv")){
        download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                      "pml-training.csv")
}

# Checking if given testing set data is locally available, else to download.
if(!file.exists("pml-testing.csv")){
        download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                      "pml-testing.csv")
}

```

### Loading and Cleaning of Data:

```{r load & clean data}

library(caret)

main_data <- read.csv("pml-training.csv")
given_testing <- read.csv("pml-testing.csv")

# Removing variables that have no Variation,as they do not help in the distinction process.
main_data <- main_data[,-nearZeroVar(main_data)]

# Removing Columns that have NA
main_data <- main_data[, colSums(is.na(main_data)) == 0]

# Removing Columns with Reading Number,Participant Name and Timestamps as they are of no use to us.
main_data <- main_data[,-(1:5)]

```
We will use the given_testing data as a final testing set (Validation).

### Creating Training and Testing Sets:

```{r data partition training / testing}

set.seed(233)

# Dividing the original Training Data (main_data) into Training and Testing sets in 7:3 ratio
inTrain <- createDataPartition(main_data$classe,p = 0.7, list = FALSE)
training <- main_data[inTrain,]
testing <- main_data[-inTrain,]

```

```{r initial plots}

library(ggplot2)
# Frequency Plot
ggplot(data = training) + geom_bar(aes(classe,fill=classe)) + labs(x = "Class", y = "Count",
                                                                   title = "Frequency of Classes")

```

### Prediction Models:

```{r reccurcive partitioning / regression tree}

set.seed(212)
library(rpart)
rpmodel <- train(classe ~ .,data = training,method = "rpart")

```

```{r regression tree plot}

library(rattle)
fancyRpartPlot(rpmodel$finalModel)

```

```{r Random forest Auto (500 Trees)}

library(rpart)
set.seed(133)
library(randomForest)
rfmodel <- train(classe ~ .,data = training,method = "rf",
                 trControl= trainControl(method = "cv",number = 5, verboseIter = FALSE),allowParallel =TRUE)

rfmodel$finalModel

```


```{r Analysing Random Forest Model}

plot(rfmodel$finalModel,main = "Error Vs Number of Trees")

```
We can clearly see here that after around 100 Trees the Error doesn't change a lot,so we create a new model with just 100 Trees,This will help us in multiple ways:

 * Helps in avoiding Over-fitting of Data
 * Helps in reduction complexity of algorithm which in turn reduces execution time of the model.

```{r Random forest Optimised (100 Trees)}

set.seed(294)
rfmodel1 <- train(classe ~ .,data = training,method = "rf",ntree=100,
                  trControl= trainControl(method = "cv",number = 5, verboseIter = FALSE),allowParallel =TRUE)

rfmodel1$finalModel

```


```{r Analysis of 500 Tree Random Forest}

starttime<- Sys.time()
predictions <- predict(rfmodel,newdata=testing)
endtime <- Sys.time()

endtime-starttime

confmatrf <- confusionMatrix(predictions,factor(testing$classe))
print(confmatrf)

five_hundred_tree <- diag(table(predictions,testing$classe))

```

```{r Analysis of 100 Tree Random Forest }

starttime<- Sys.time()
predictions <- predict(rfmodel1,newdata=testing)
endtime <- Sys.time()

endtime-starttime
confmatrf1 <- confusionMatrix(predictions,factor(testing$classe))
print(confmatrf1)
one_hundred_tree <- diag(table(predictions,testing$classe))

```

As we see Both the models (500 Trees and 100 Tress) perform very similarly in terms of accuracy but the execution time of the second model (100 Trees) is about 1/3th of that of the Original model (500 Trees).

```{r comparison}

new_data <- data.frame(one_hundred_tree,five_hundred_tree,
                       as.numeric(table(testing$classe)))

colnames(new_data) <- c("One Hundred Trees","Five Hundred Trees",
                        "Actual Class Distribution")

knitr::kable(new_data,caption = "Comparison of performance of 500 vs 100 Trees")

```
We will go along with rfmodel1 (100 Tree Model) as it performs as well as rfmodel (500 Tree model) and considerably better than the Regression Tree model.

### Applying Model to predict given Test dataset.

```{r}

predict_given <- predict(rfmodel1,newdata = given_testing)
print(predict_given)

```



